---
title: 'Assignment 4: Tidy Data'
author: "Kyle Wlodarczyk"
date: "October 4, 2019"
output: 
  html_document:
    keep_md: true
    df_print: paged
    toc: yes
    theme: cerulean
---

Before we begin, let us load some packages that we will potentially be using. For the first two exercises of this assignment, we will be wrangling part of the gapminder dataset before wrangling an independent dataset for the third exercise.

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(gapminder)
library(tibble)
library(DT)
```

# Exercise 1 - Univariate Option 1

Let's take a look at the life expectancy of two countries, such as Poland and Switzerland, for each year. Note that in the table below, the values have been rounded and are given in the unit "years".

```{r, message = FALSE, warning = FALSE}
gapminder_r = gapminder
gapminder_r$lifeExp = round(gapminder_r$lifeExp, 1)

Ex1 <- gapminder_r %>%
  select(country, year, lifeExp) %>%
  filter(country == "Poland" | country == "Switzerland") %>%
  pivot_wider(id_cols = year,
              names_from = country,
              values_from = lifeExp)

datatable(Ex1)
```

To get a better idea of how the life expectancies of the two countries are changing relative to each other, let's graph them against each other since visuals are powerful communication tools.

```{r, message = FALSE, warning = FALSE}
Ex1 %>%
  ggplot() +
  geom_point(aes(Poland, Switzerland)) +
  xlab("Poland") +
  ylab("Switzerland") +
  ggtitle("Life Expectancy in Poland and Switzerland") +
  geom_text(aes(Poland, Switzerland, label = year), hjust = .1, vjust = -.3) +
  theme_bw()
```

We can see that the life expectancy in Poland stagnates between 1972 and 1992 while it continues to increase in Switzerland.

Now let's return the data back to the format found in the gapminder dataset, where observations are listed for each country and arranged by country followed by year.

```{r, message = FALSE, warning = FALSE}
Ex1 %>%
  pivot_longer(cols = c(Poland, Switzerland),
               names_to = "country",
               values_to = "lifeExp") %>%
  select(country, year, lifeExp) %>%
  arrange(country) %>%
  datatable()
```


# Exercise 2 - Multivariate Option 1

After reviewing the life expectancy of Poland and Switzerland for each year, let's see how the GDP per capita changes with life expectancy. Note that in the table below, the values have been rounded and are given in the unit "years" for life expectancy or "dollars" for GDP per capita.

```{r, message = FALSE, warning = FALSE}
gapminder_r$gdpPercap = round(gapminder_r$gdpPercap, 2)

Ex2 <- gapminder_r %>%
  select(country, year, lifeExp, gdpPercap) %>%
  filter(country == "Poland" | country == "Switzerland") %>%
  pivot_wider(id_cols = year,
              names_from = country,
              values_from = c(lifeExp, gdpPercap)) %>%
  select(year, lifeExp_Poland, gdpPercap_Poland, lifeExp_Switzerland, gdpPercap_Switzerland)

datatable(Ex2)
```

Now that we've seen how GDP per capita compares to life expectancy for both Poland and Switzerland, let's return the data back to the format found in the gapminder dataset, where observations are listed for each country and arranged by country followed by year.

```{r, message = FALSE, warning = FALSE}
Ex2 %>%
  pivot_longer(cols      = -year, 
               names_to  = c(".value", "country"),
               names_sep = "_") %>%
  select(country, year, lifeExp, gdpPercap) %>%
  arrange(country) %>%
  datatable()
```


#Exercise 3

Let us now switch gears and look at a wedding guestlist and email addresses.

```{r, message = FALSE, warning = FALSE}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
```

## Exercise 3.1

It is convenient to have all of the information that we need in one tibble instead of referencing two tibbles. Let's add each guest's email address (if available) to the guestlist.

```{r, message = FALSE, warning = FALSE}
email_sep <- email %>%
  separate_rows(guest, sep = ", ") %>%
  rename(name = guest)
guest %>%
  select(-party) %>%
  left_join(email_sep, by = name)
```

## Exercise 3.2

We have more emails than people on the guestlist; the people below are not on the guestlist, yet we still have their emails. 

```{r, message = FALSE, warning = FALSE}
email_sep %>%
  anti_join(guest, by = name)
```

## Exercise 3.3

Let's add those people who we have emails for but aren't on the guestlist (seen in Exercise 3.2 above) to the guestlist.

```{r, message = FALSE, warning = FALSE}
guest %>%
  full_join(email_sep, by = name)
```

